White Wine Quality Analysis by Navpreet Singh
========================================================

#### This report explores the data set of white wines of different quality and 
chemical composition. The data set contains 4898 observations of 12 variables. 
This report contains Univariate, Bivariate and Multivariate analysis of 
different variables of the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine_data <- read.csv('wineQualityWhites.csv')
wine_data <- subset(wine_data, select=-c(X))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, dimensions}
dim(wine_data)
```


> This report explores a dataset containing different quality and chemical 
composition of over 4800 wines.


```{r echo=FALSE, message=FALSE, warning=FALSE, cutting_and_factoring}
# Dividing different ratings of wines into 3 buckets[0-5, 6-7, 8-10]
wine_data$quality.bucket <- with(wine_data, cut(quality, breaks=c(0, 5, 7, 10)))

# Converting Quality into a factor
wine_data$quality <- factor(wine_data$quality)

str(wine_data)
```

- Since quality is a discrete variable having values between 3 and 9, converted 
it to a factor categorical variable.
- Created a new categorical variable quality.bucket using cut method by 
categorizing different quality scores as **(0,5] - Poor**, **(5,7] - Average** 
and **(7,10] - Good**.

```{r echo=FALSE, message=FALSE, warning=FALSE, table_on_quality}
table(wine_data$quality)
```

- Most of the wines have a score of 5 and 6. Only a small number of wines 
are very good quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, summary}
summary(wine_data)
```

------

# Univariate Plots Section

### Quality Distributions

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Distributions}
ggplot(aes(x = quality), data = wine_data) +
  geom_bar()
```

In this plot we can see the distribution of Quality of wines, Lowest scoring 
wines have a score of 3 while highest score is 9 for some wines. Most of the 
wines falls under the score of 5 to 7.

### Distribution of quality as Good, Poor and Average

```{r echo=FALSE, message=FALSE, warning=FALSE, Good_Poor_Average_quality}
ggplot(aes(x = quality.bucket), data = wine_data) +
  geom_bar()
```

This histogram divide the wines in **Poor**, **Average** and **Good** category. 
Most of the wines are under average catogory.

### Distribution of all quantative variables

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
wine_data[,c(1:11)] %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = 'free') +
  geom_histogram()
```

The plot above gives the quick glance at the distribution of all quantitative 
variables present in the data. Below is the overview of some of the important 
variables:

- Alcohol distribution is very dispersed between all the values.\
- Sulphates shows the signs on bimodel distribution.\
- pH and total sulfurdioxide looks evenly spreaded and normally distributed.\
- Distribution of Chlorides is very less dispersed.\

### Alcohol distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_distribution}
p_alcohol <- ggplot(aes(x=alcohol), data=wine_data) +
  geom_histogram()

p_alcohol_box <- ggplot(aes(x='',y=alcohol), data=wine_data) +
  geom_boxplot()

grid.arrange(p_alcohol, p_alcohol_box, nrow=1)
```

As we can see from the boxplot, the varience in the alcohol quantity is large, 
with middle 50% of wines contains an alcohol percentage in the range of 
9.5% to 11.5%

### pH Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, pH_distribution}
p_pH <- ggplot(aes(x = pH), data = wine_data) +
  geom_histogram()

p_pH_box <- ggplot(aes(x='',y=pH), data=wine_data) +
  geom_boxplot()

grid.arrange(p_pH, p_pH_box, nrow=1)
```

The pH distribution is almost normal with small varience. But it contains a 
number of outliers. Most of the wines are in the pH range of 3-3.3

### Residual Sugar Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Residual_Sugar_Distribution}
p_sugar <- ggplot(aes(x=residual.sugar), data=wine_data) +
    geom_histogram()

p_sugar_log <- ggplot(aes(x=residual.sugar), data=wine_data) +
    scale_x_log10() +
    geom_histogram()

grid.arrange(p_sugar, p_sugar_log, nrow=1)
```

The distribution of residual.sugar is rightly skewed. By taking a log10 of the 
distribution transformed to be bimodal.

### Volatile Acidity Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Volatile_Acidity, fig.height=8, fig.width=6}
p1 <- ggplot(aes(x = volatile.acidity), data = wine_data) +
  geom_histogram(binwidth = 0.01)

p2 <- ggplot(aes(x = volatile.acidity, color=quality.bucket), data=wine_data) + 
  geom_freqpoly() +
  scale_color_manual(values=c("purple", "blue", "brown"))

grid.arrange(p1, p2, nrow=2)
```

- The volatile acidity distribution is slightly right skewed.
- The volatile acidity of different wine qualities are in second plot shown 
above. It is clear from the distribution that high quality wines have less 
volatile acidity content

### Density Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_distribution}
p1 <- ggplot(aes(x=density), data=wine_data) +
  geom_histogram()

p2 <- ggplot(aes(x=density), data=wine_data) +
  geom_histogram() +
  scale_x_continuous(limits=c(quantile(wine_data$density, 0.01), 
                              quantile(wine_data$density, 0.99)),
                     breaks = seq(0.99, 1.04, 0.01))

grid.arrange(p1, p2, nrow=1)
```

Most of the values (98 %) are distributed in the range of 0.99-1.00 g/cm^3^. The
density variable also have some outliers which are stretching the tail of 
distribution to 1.04 g/cm^3^.

### Total Sulfur Dioxide Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Total_Sulfur_Dioxide, fig.width=8, fig.height=10}
p1 <- ggplot(aes(x=total.sulfur.dioxide), data=wine_data) +
  geom_density() 

p2 <- ggplot(aes(x=total.sulfur.dioxide), data=wine_data) +
  geom_density() +
  scale_x_sqrt()

grid.arrange(p1, p2, ncol=1)
```

- The first plot shows the distrubution of total sulfur dioxide. The 
distribution has a small tail. Most of the total 
sulfur dioxide seems to a have a spread between 70 mg/dm^3^ and 210 mg/dm^3^.
- The second plot represents the square root transformation of the distribution 
of total sulfur dioxide. It eliminated the tail and the values are distributed 
normally.

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains 4898 observations of 12 variables. I have created one more 
variable quality.bucket which categorized the wine as "Good", "Average" or 
"Poor" based on its quality score.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interests that I observed are density, alcohol and quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think volatile acidity, Residual Sugar, pH and all different type of supfur 
dioxide features will help in the investigation.

### Did you create any new variables from existing variables in the dataset?

I have created one extra feature quality.bucket which categorized the wine as 
"Good", "Average" or "Poor" based on its quality score.

------

# Bivariate Plots Section

### Relationship between different variables

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots, fig.height=25, fig.width=25}
ggpairs(data=subset(wine_data, select=-c(quality.bucket)), 
        lower = list(
            continuous = "smooth",
            combo = "facetdensity"
          )
        )

```

The above plot shows the relation of each variable present in dataset with every 
other variable present. We will explore some of them in more details.

### Alcohol and Quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_quanity}
ggplot(aes(x=alcohol), data=wine_data) +
  geom_density(aes(fill=quality.bucket), alpha=0.5) +
  theme_dark() +
  scale_fill_brewer(labels = c("Poor", "Average", "Good"))
```

As we can see from the plot above, More quantity of alcohol is one of the major 
driving force in the quality of wines.

### Free Sulfur Dioxide VS Total Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, Free_vs_Total}
ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide), data=wine_data) +
  geom_point(alpha=1/4, position = position_jitter(h=0), color='orange') +
  geom_smooth(method='lm') +
  scale_x_continuous(limits=c(0,150), breaks=seq(0,150,20)) +
  scale_y_continuous(limits=c(0,400), breaks=seq(0,400,40))
```

The plot shows a strong relationship between free Sulfur dioxide and total 
sulfur dioxide. This is because of the reason that free sulfur dioxide is also 
a part of total sulfur dioxide.

### Acidity and pH:
```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity_pH}
ggplot(aes(x=fixed.acidity, y=pH), data=wine_data) +
  geom_point(position = position_jitter(h=0), color='brown') +
  geom_smooth(method='lm', aes(color=quality.bucket)) +
  scale_x_continuous(limits=c(quantile(wine_data$fixed.acidity, 0.01), 
                              quantile(wine_data$fixed.acidity, .99))) +
  scale_y_continuous(limits=c(quantile(wine_data$pH, 0.01), 
                              quantile(wine_data$pH, .99)))
```

The fixed acidity is having a strong negative relationship with pH for all 3 
quality buckets. This is very obvious behaviour since pH and acidity are 
inversly proportional to each other.

### Density VS Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol}
ggplot(aes(x=density, y=alcohol), data=wine_data) +
  geom_point(position = position_jitter(h=0), alpha=1/4, color='blue')  +
  geom_smooth(method='lm', aes(color='red')) +
  scale_x_continuous(limits=c(0.98, 1.02), breaks=seq(0.98, 1.02, 0.01)) +
  scale_y_continuous(limits=c(8,15), breaks=seq(8, 15, 1))
```

The plot shows a negative relation between density and alcohol. Since, increased
 density of water make alcohol more diluted and its concentration will be less.

### Density VS Residual Sugar
```{r echo=FALSE, message=FALSE, warning=FALSE, Density_sugar}
ggplot(aes(x=density, y=residual.sugar), data=wine_data) +
  geom_point(position = position_jitter(h=0), alpha=1/2)  +
  geom_smooth(method='lm', color='red') +
  scale_x_continuous(limits=c(0.99, 1.01), breaks=seq(0.99, 1.01, 0.01)) +
  scale_y_continuous(limits=c(0, 40), breaks=seq(0, 40, 10))
```

The plot above is showing a positive relationship with most values in the range 
of 0.99-1 g/cm^3^ for density and 0-15 g/dm^3^ for residual sugar.

### Citric Acid VS Fixed Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Citric_fixedAcidity}
ggplot(aes(x=citric.acid, y=fixed.acidity), data=wine_data) +
  geom_point(position = position_jitter(h=0), color='blue')  +
  scale_x_continuous(limits=c(0, 1.0), breaks=seq(0, 1.0, 0.2)) +
  scale_y_continuous(limits=c(0, 12), breaks=seq(0, 12, 2))
```

Citric Acid and Fixed Acidity does not show very strong relationship. Most of 
the values are clustered in 0.2-0.4 g/dm^3^ for citric acid and 5-9 g/dm^3^ for 
fixed acidity.


### Mean and Median of Alcohol grouped by quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Mean_Median_alcohol}
wine_by_quality <- wine_data %>%
  group_by(quality) %>%
  summarise(mean_alcohol = mean(alcohol),
            median_alcohol = median(as.numeric(alcohol)),
            n = n()) %>%
  arrange(quality)

wine_by_quality
```

### Mean of Alcohol by Quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Mean_by_Quality}
ggplot(aes(x=quality, y=mean_alcohol, group=1), data=wine_by_quality) +
  #geom_line(aes(y=median_alcohol), color='blue') +
  geom_line(color='red')
```

The mean values of alcohol content surprisingly shows a dip at a quality level 
of 5 but again increased with increasing quality levels.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

- Quality of wine is majorly depends upon the alcohol content in the wine. There 
is a positive relation between alcohol and quality of wine. It can be used to 
predict the quality of wine.
- Density of water shows a negative relation with alcohol as more water make 
alcohol more diluted making less alcohol content present in the wine and affect 
its quality.
- Acidity is having a strong negative relationship with pH. This is very 
obvious behaviour since pH and acidity are inversly proportional to each other.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

- Density and Residual sugar exhibit strong positive relationship which was 
surprising for me. I was assuming that with increasing density, sugar level 
should come down.
- I was hoping to see a strong positive relationship between citric acid and 
fixed acidity. But the relation was quite weak.

### What was the strongest relationship you found?

- Free sulfur dioxide and Total sulfur dioxide shows a strong positive 
relationship.
- Alcohol and Quality has strong positive relationship.
- Density and Residual sugar has strong positve relationship.

- Alcohol and Density shows a strong negative relationship.
- Acidity and pH value shows a negative relationship.

------

# Multivariate Plots Section

### Alcohol VS Density VS Quality Buckets

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_1, fig.height=8, fig.width=12}
ggplot(aes(x=alcohol,y=density, color=quality.bucket), data=wine_data) +
  geom_jitter(alpha=1/2, size=5) +
  theme_dark() +
  scale_x_continuous(limits=c(quantile(wine_data$alcohol, 0.025), 
                              quantile(wine_data$alcohol, .975))) +
  scale_y_continuous(limits=c(quantile(wine_data$density, 0.025), 
                              quantile(wine_data$density, .975))) +
  scale_color_brewer(palette = 'RdPu')
```

The plot above display middle 95% of density and alcohol data. I have separated 
the data on the bases of quality buckets. As we can see from the plot, the 
quality of wine increasing as the alcohol content is increasing and density of 
water is decresing.

### Fixed Acidity VS pH VS Quality Buckets

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_2, fig.height=8, fig.width=12}
ggplot(aes(x=fixed.acidity, y=pH, color=quality.bucket), data=wine_data) +
  geom_point(position = position_jitter(h=0), alpha=1/2, size=4) +
  geom_smooth(method='lm', aes(color=quality.bucket), se=F) +
  theme_dark() +
  scale_x_continuous(limits=c(quantile(wine_data$fixed.acidity, 0.01), 
                              quantile(wine_data$fixed.acidity, .99))) +
  scale_y_continuous(limits=c(quantile(wine_data$pH, 0.01), 
                              quantile(wine_data$pH, .99))) +
  scale_color_brewer(type='qual', palette = 3)
```

The plot above is among fixed acidity, pH and quality of wines. pH and acidity 
have strong negative relationship with each other. The good quality wines have 
slightly higher levels of pH for the same acidity levels as compare to low 
quality wines. It means good wines are less acidic in nature.


### Free Sulfur Dioxide VS Total Sulfur Dioxide VS Quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_3, fig.height=8, fig.width=12}
ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide), data=wine_data) +
  geom_point(alpha=1/2, position = position_jitter(h=0), aes(color=quality), 
             size=4) +
  geom_smooth(method='lm', aes(color=quality), se=FALSE) +
  theme_dark() +
  scale_x_continuous(limits=c(quantile(wine_data$free.sulfur.dioxide, 0.01), 
                              quantile(wine_data$free.sulfur.dioxide, .99))) +
  scale_y_continuous(limits=c(quantile(wine_data$total.sulfur.dioxide, 0.01), 
                              quantile(wine_data$total.sulfur.dioxide, .99))) +
  scale_color_brewer()
```

The above diagram shows 99% middle of free and total sulfur dioxide data. As we 
can see from the diagram above that for high quality wines, the increase in 
total sulfur dioxide values is very low as compare to low quality wines.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

- First relationship I observed is among alcohol, water density and quality of 
wines. The quality of wines is increasing as the alcohol content is increasing 
and density of water is decreasing.
- Second relationship I observed is among fixed acidity, pH and quality of wines. 
pH and acidity have strong negative relationship with each other, but the 
point to noted here is that good quality wines have slightly higher levels of pH 
for the same acidity levels as compare to low quality wines. It means good 
wines are less acidic in nature.
-Third relationship I observed is among free sulfur dioxide, total sulfur 
dioxide and quality of wines. For high quality(especially score 9) wines, the 
total sulfur dioxide content is incresing at slower rate as compare to wines 
with low quality.

### Were there any interesting or surprising interactions between features?

The first relationship among alcohol, water density and quality of wines is 
interesting one. highly alcoholic wines are termed as good quality wines. People 
like to drink wines which have highly concentrated alcoholic content which can 
give them easy high.

------

# Final Plots and Summary

### Plot One - Density distribution by Quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One, fig.height=8, fig.width=12}
ggplot(aes(x=quality, y=density), data=wine_data) +
  geom_boxplot() +
  scale_y_continuous(limits=c(quantile(wine_data$density, 0.01), 
                              quantile(wine_data$density, 0.99)),
                     breaks = seq(0.99, 1.0, 0.005)) +
  labs(title = "\nDistribution of Density by Quality\n", x = "\nQuality", 
       y = "Density (g / cm^3)\n")
```

### Description One

This is a boxplot of density of water for different qualities of wine. This plot 
shows the negative influence of water density on the quantity of wine. As with 
decreasing density, quality of wine is incresing.

### Plot Two - Alcohol by Quality Buckets
```{r echo=FALSE, Plot_Two, fig.height=6, fig.width=10}
ggplot(aes(x=alcohol), data=wine_data) +
  geom_density(aes(fill=quality.bucket), alpha=0.5) +
  scale_x_continuous(limits=c(8, 15), breaks = seq(8, 15, 1)) +
  scale_y_continuous(limits=c(0, 0.7), breaks = seq(0, 0.7, 0.1)) +
  theme_dark() +
  labs(title = "\nAlcohol Density per Quality Bucket\n", 
       x = "\nAlcohol (% by Volume)", y = "Density (mg / l)\n", 
       fill = "Quality Bucket\n") +
  scale_fill_brewer(labels = c("Poor", "Average", "Good"))
```

### Description Two
This is the plot of alcohol density per quality bucket. It shows that more 
quantity of alcohol is one of the major driving force in the quality of wines.

### Plot Three - Acidity VS pH

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three, fig.height=8, fig.width=12}
ggplot(aes(x=fixed.acidity, y=pH, color=quality.bucket), data=wine_data) +
  geom_point(position = position_jitter(h=0), alpha=1/2, size=4) +
  geom_smooth(method='lm', aes(color=quality.bucket), se=F) +
  scale_x_continuous(limits=c(quantile(wine_data$fixed.acidity, 0.01), 
                              quantile(wine_data$fixed.acidity, .99)),
                     breaks = seq(5, 9, 0.5)) +
  scale_y_continuous(limits=c(quantile(wine_data$pH, 0.01), 
                              quantile(wine_data$pH, .99)),
                     breaks = seq(2.9, 3.6, 0.1)) +
  theme_dark() +
  labs(title="\nAcidity vs pH\n", x="\nFixed Acidity (g / dm^3)", y="pH\n", 
       color="Quality Bucket\n") +
  scale_color_brewer(type = 'seq', labels = c("Poor", "Average", "Good"))
```

### Description Three

This is the plot among fixed acidity, pH and quality of wines. pH and acidity 
have strong negative relationship with each other. Good quality wines have 
slightly higher levels of pH as for the same acidity levels as compare to low 
quality wines. It means good wines are less acidic in nature.


------

# Reflection

In this report, exploratory data analysis was performed on a data set of white 
wines of different quality and chemical composition. The data set contains 
4898 observations of 12 variables.

The analysis started with univariate analysis where single varibale was analyzed. 
Followed by bivariate analysis where the relationship between two variables was 
analyzed. It ended with multivariate analyses where relation between more than 2 
variables have been analyzed.

Following are the key points of the analysis performed:

- The amount of alcohol in the wine is one of the major parameter in the quality 
of wine. with wines having more alcohol got good score in wine quality test as 
compare to wines with less alcohol.
- The quantity of water density in the wine also plays a key role. It bears a 
negative relationship with the quality of wine. With the increase in water 
density, The qualiity of wine decreases.
- As with various acids present in wine, I observed that as the quantity of 
acids increases, the pH value is getting decreases as so the quality of wines. 
The effect is not very high still to be noted.

**Struggles and Difficulties** : I don not have much prior knowledge about the 
wine industry and how the quality of wine is tested. During this analysis I came 
to know about various components(chemicals) which are used in wine making 
process. Knowing how these different chemicals affect wine quality was major 
chellenge for me.

**Success** : While drawing Univariate and Bivariate plots, I came to know about 
the affect of some of the chemicals (e.g. alcohol, pH, density, acidity) on wine 
quality. While drwaing these plots I came to know more about data and their 
relationships.

The analysis can be carry forward in coming future with more latest data since 
present data in from year 2009. Also, this data contains varients of only
Portuguese "Vinho Verde" wine. It will be much better if other wines samples 
also include in the dataset to make it more diversify.


------

